---
title: 'PSPG Bootcamp 2020: Data Manipulation and Plotting in R'
author: "Vincent Chang and Jackie Ernest"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 1
  pdf_document: default

---

<b>Objectives: </b><br>
<b>1) Learn the basics of dplyr </b> <br>
    -Selecting, Filtering, and Arranging <br>
    -Making a series of function with pipes %>% <br>
    -Mutate new and existing columns <br>
    -Wide and long tables <br>
<br>
<b> 2) Learn the basics of ggplot2 </b> <br>
    -Setting up your dataframe for easy plotting <br>
    -Plot lines, scatter plots, boxplots <br>
    -Compare different subpopulations <br>
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Installing and Loading packages
```{r warning=F, message=F}
#Comment out if you've already installed them
#install.packages("dplyr")
#install.packages("tidyr")
#install.packages("ggplot2")

#Load libraries
library(dplyr)
library(tidyr)
library(ggplot2)
```




# Loading your dataset
```{r}
#setwd("~/Documents/Courses/Bootcamp/bootcamp_2020/")
PKdata <- read.csv("simtab.csv")

#Let's take a glimpse at your dataset
head(PKdata)
#Alternatively, click on PKdata in your Global Environment toolbar on the right

#Let's open up the data dictionary too, it's always good practice to know your dataset
```

## Understanding your dataset
```{r}
#Below are a few other ways to get a high-level glance at your data
glimpse(PKdata)
summary(PKdata)

#Don't know what a function does? Add a question mark to view its documentation
?table

#You can use table to compare counts of different variables
table(PKdata$SEX, PKdata$HIV)

#To add titles
table("SEX" = PKdata$SEX, "HIV" = PKdata$HIV)

#How many females were in each dose level?
table("DOSE" = PKdata$DOSE,"SEX" = PKdata$SEX)

```

Dplyr Cheatsheet
https://raw.githubusercontent.com/rstudio/cheatsheets/master/data-transformation.pdf

# Learn the basics of dplyr
## Selecting
```{r}
#You can select specific columns using select()
#Let's de-identify our dataset by selecting the columns we want
deidentified <- select(PKdata, ID, TIME, DV, DOSE, HT, BW, SEX, HIV)
head(deidentified)

#You can also do this by "subtracting" the columns you don't want with -
deidentified <- select(PKdata, -PT_BIRTHDATE, -PT_ADDRESS, -PT_USUBJID)
head(deidentified)

#You can also search through your columns with helpers: contains(), starts_with(), ends_with()
What_did_this_do <- select(PKdata, ends_with("V"))
head(What_did_this_do)
#Write your answer commented out below:
#removed all columns that end with V

#Can you use helpers to create the same deidentified dataset?
deidentified <- select(PKdata, -starts_with("PT"))
head(deidentified)


```

## Filtering
```{r}
#You can filter out specific rows using filter()
#For example, selecting all patients who received a specific dose
Dosed_600 <- filter(PKdata, DOSE == 600)
head(Dosed_600)

#You can use any logical function in R to designate your filter criteria
# > greater than, <= greater equal than
# < less than, <= lesser equal than
# ! not
# & and
# | or
#Can you filter for patients with body weight greater or equal than 45kg?
WT.GT.45 <- filter(PKdata, BW >= 45)
head(WT.GT.45)

#You can also use & or | to create more complex logical functions
#Let's filter for patients that received a dose greater than or equal to 600 and HIV+
Dosed600up_HIV <- filter(PKdata, DOSE >= 600 & HIV == 1)
head(Dosed600up_HIV)

#Challenge: Filter for patients who received the lowest or highest dose, and is female with HIV or male
#Remember there are always many ways to code something :)
Challenge <- filter(PKdata, DOSE != 600 & (HIV == 1 | SEX == 1))
head(Challenge)

```

## Arranging
```{r}
#arrange() will reorder rows or columns according to your specifications (default from low to high)
#For example in PK data we will often organize our rows by patient ID, then by time.
PKdata <- arrange(PKdata, ID, TIME)
```

## Putting it all together with Pipes %>%
```{r}
#Pipes are used to connect the output of one function to the input of another
#Let's deidentify our dataset, filter for 600mg dose, and arrange by id and time all together
Dose_600 <- PKdata %>%
  select( -starts_with("PT")) %>%
  filter(DOSE == 600) %>%
  arrange(ID, TIME)

#Try this one for yourself!
#Let's try filtering for HIV+ or body weight less than 45kg, then...
#we want to see the trough levels of each patient (defined as the 24hr timepoint)  

Trough <-   PKdata %>%
  select( -starts_with("PT")) %>%
  filter(HIV == 1 | BW < 45) %>%
  filter(TIME == 24)

View(Trough)
  
  
```

## Mutate
```{r}
#mutate() can create new columns or alter existing ones
#For example, BMI is often calculated from HT and WT
#BMI = weight(kg)/height(m)^2

PKdata <- PKdata %>%
  mutate(BMI = BW/(HT/39.37)^2) #are the units correct? check the data dictionary!

#Now you try making a new column and converting DV from mg/L into micromoles!
#The drug used here is remdesivir

PKdata<- PKdata %>%
  mutate(Conc_uM = DV/1000/602.6*1000000) #mg/L *1g/1000mg * 1mol/602.6g * 1000000uM/1M


```

# Learn the basics of ggplot2

## Our first Plot!!
ggplot2 cheatsheet: https://raw.githubusercontent.com/rstudio/cheatsheets/master/data-visualization-2.1.pdf
```{r}
ggplot(PKdata) + 
  geom_point(aes(x=TIME, y=DV)) ##What are we trying to plot with a PK profile?


#Do you like lines better?
ggplot(PKdata) + 
 geom_line(aes(x=TIME, y=DV,group=factor(ID)),alpha = 0.1, linetype=2,size=0.1)

#Let's put them on the same graph with a +

ggplot(PKdata) + 
 geom_line(aes(x=TIME, y=DV,group=factor(ID)),alpha = 0.1, linetype=2,size=0.1) +
  geom_point(aes(x=TIME, y=DV))

```

## Graphing Subpopulations
```{r}
ggplot(PKdata) + 
  geom_point(aes(x=TIME, y=DV, color = as.factor(DOSE))) + 
  geom_line(aes(x=TIME, y=DV, group=factor(ID), color = as.factor(DOSE)),alpha = 0.1, linetype=2,size=0.1)
```

## Another way to stratify
```{r}
#It's still a little too cluttered
ggplot(PKdata) + 
  geom_point(aes(x=TIME, y=DV, color = as.factor(DOSE))) + 
  geom_line(aes(x=TIME, y=DV, group=factor(ID), color = as.factor(DOSE)),alpha = 0.1, linetype=2,size=0.1) +
  facet_wrap(.~DOSE)
```

## Let's try some on your own
```{r}
#Copy and paste from above, and let's plot stratified by HIV status
ggplot(PKdata) + 
  geom_point(aes(x=TIME, y=DV, color = as.factor(HIV))) +
  geom_line(aes(x=TIME, y=DV, group=factor(ID), color = as.factor(HIV)),alpha = 0.1, linetype=2,size=0.1) +
  facet_wrap(.~DOSE)
```

## Two stratifications
```{r}
#Now let's try adding 2 stratifications: color by HIV and facet_wrap by Dose~SEX. Then try the other way around!
ggplot(PKdata) + 
  geom_point(aes(x=TIME, y=DV, color = as.factor(HIV))) + 
  geom_line(aes(x=TIME, y=DV, group=factor(ID), color = as.factor(HIV)),alpha = 0.1, linetype=2,size=0.1) +
  facet_wrap(DOSE~SEX)

ggplot(PKdata) + 
  geom_point(aes(x=TIME, y=DV, color = as.factor(SEX))) +
  geom_line(aes(x=TIME, y=DV, group=factor(ID), color = as.factor(SEX)),alpha = 0.1, linetype=2,size=0.1) +
  facet_wrap(DOSE~~HIV)

#Challenge: Plot stratified by body weight, one color for above 45kg and another color for below. Make sure to have a legend :)
#Hint: you will need to manipulate your dataframe

PKdata <- PKdata %>%
  mutate(WT_FLAG = ifelse(BW > 45, 1, 0))

ggplot(PKdata) + 
  geom_point(aes(x=TIME, y=DV, color = as.factor(WT_FLAG))) + 
  geom_line(aes(x=TIME, y=DV, group=factor(ID), color = as.factor(WT_FLAG)),alpha = 0.1, linetype=2,size=0.1) +
  facet_wrap(.~DOSE)



```

# Summarise
## Dplyr can be used for simple summary statistics
```{r}
#You can use summarise to calculate means, ranges, standard deviations, quartiles, etc.
Summary_stats <- PKdata %>%
  summarise(c_mean = mean(DV),
            c_stdev = sd(DV),
            BW_mean = mean(BW),
            BW_upper = quantile(BW, probs = 0.975),
            BW_lower = quantile(BW, probs = 0.225))

View(Summary_stats)

```

## Combined with group_by(), summarise is a force to be reckoned with..
```{r}
#Let's calculate subpopulation summary statistics
#For example, let's calculate body weight summary statistics by SEX
BW_by_SEX <- PKdata %>%
  group_by(SEX) %>%
  summarise(BW_mean = mean(BW),
            BW_upper = quantile(BW, probs = 0.975),
            BW_lower = quantile(BW, probs = 0.225))

#Let's plot it!
ggplot(BW_by_SEX) +
  geom_point(aes(x=SEX, y=BW_mean)) +
  geom_errorbar((aes(x=SEX, ymin=BW_lower, ymax=BW_upper)))


#An easier way to do it in just ggplot, there are always multiple ways to do the same thing
ggplot(PKdata) +
  geom_boxplot(aes(x=as.factor(SEX), y=BW))
```

## Let's calculate the median PK profile
```{r}
#Let's start with calculating median and percentile profiles 
Med <- PKdata %>%
  group_by(TIME) %>%
  summarise(median = median(DV),
            upper = quantile(DV, probs=0.975), #Upper 95th percentile
            lower = quantile(DV, probs=0.225)) #Lower 95th percentile

#Let's plot it
ggplot(Med) +
  geom_line(aes(x=TIME, y=median)) +
  geom_ribbon(aes(x=TIME, ymin=lower, ymax=upper), alpha=0.3)
```

## Let's compare median PK profiles of subpopulations
```{r}
#Let's do the same thing but subset by DOSE
DOSE_Med <- PKdata %>%
  group_by(DOSE,TIME) %>%
  summarise(median = median(DV),
            upper = quantile(DV, probs=0.975), #Upper 95th percentile
            lower = quantile(DV, probs=0.225)) #Lower 95th percentile

#Let's plot it
ggplot(DOSE_Med) +
  geom_line(aes(x=TIME, y=median, color = as.factor(DOSE))) +
  geom_ribbon(aes(x=TIME, ymin=lower, ymax=upper, fill = as.factor(DOSE)), alpha=0.3)
```

## Let's try another one...
```{r}
#Median PK profiles, subset by HIV

HIV_Med <- PKdata %>%
  group_by(HIV,TIME) %>%
  summarise(median = median(DV),
            upper = quantile(DV, probs=0.975), #Upper 95th percentile
            lower = quantile(DV, probs=0.225)) #Lower 95th percentile

#Let's plot it
ggplot(HIV_Med) +
  geom_line(aes(x=TIME, y=median, color = as.factor(HIV))) +
  geom_ribbon(aes(x=TIME, ymin=lower, ymax=upper, fill = as.factor(HIV)), alpha=0.3)


```

## This doesn't really work...the different dosing levels make the stratification cluttered
```{r}
#We need to subset by DOSE and HIV
#Prepare the dataset yourself

HIV_Med <- PKdata %>%
  group_by(DOSE, HIV,TIME) %>%
  summarise(median = median(DV),
            upper = quantile(DV, probs=0.975), #Upper 95th percentile
            lower = quantile(DV, probs=0.225)) #Lower 95th percentile

#Let's plot it
ggplot(HIV_Med) +
  geom_line(aes(x=TIME, y=median, color = as.factor(HIV))) +
  geom_ribbon(aes(x=TIME, ymin=lower, ymax=upper, fill = as.factor(HIV)), alpha=0.3) +
  facet_wrap(.~DOSE) +
  scale_y_log10()

```

